---
title: "The `zoon` R package for reproducible and shareable species distribution modelling"
author: "Nick Golding, Tom August, Tim C.D. Lucas, Emiel Van Loon & Greg McInerny"
bibliography: zoon_app.bib
csl: mee.csl
output:
  pdf_document:
    fig_caption: yes
    includes:  
      in_header: latex_preamble.tex
---

# Abstract

1. The diverse array of software and methodological research available for species distribution modelling (SDM) hinders independent evaluation of new methods and their dissemination to SDM users.  

2. The `zoon` R package encodes SDM analyses as a simple, but fully reproducible workflow of five steps: obtaining occurrence data, obtaining covariate data, pre-processing these data, fitting a model, and generating outputs.

3. Each of these steps is carried out by one or more community-contributed software modules maintained in a version-controlled online repository and built upon existing SDM code form R-packages and the research community.

4. `zoon` workflows are re-runnable records of the data, code and results of an entire SDM analysis and can be easily reproduced, scrutinized and extended by the whole research community.

5. We demonstrate `zoon` by recreating SDM analyses from two published research articles as zoon workflows, which readers can interrogate and extend.


```{r knitrOpts, echo = FALSE, cache = FALSE, eval = TRUE}
# set up knitr options
knitr::opts_chunk$set(fig.path = 'figs/',
               message = FALSE,
               warning = FALSE,
               dev = c('png'),
               cache = TRUE)
```

```{r raster_dir, echo = FALSE, cache = FALSE, eval = TRUE}
# set a place for the rasters to be downloaded to (to stop them being grabbed
# again each time)
raster_dir <- './raster_data'
if (!dir.exists(raster_dir)) dir.create(raster_dir)
options("rasterDataDir" = raster_dir)
```

# Introduction

Like many areas of quantitative science Species Distribution Modelling (SDM) has grown rapidly [@Barbosa2015] fueled by the availability of diverse modelling software and data.
Over the last 20 years, SDM has developed into a very large research community and vast literature, and is now one of “the most widely-reviewed topics in the ecological literature” (Araújo & Peterson XXXX).
Most SDM software, and that enabling other other computational sub-disciplines, focus on  analytical tasks (such as modelling and statistics) rather than enabling scientists to produce reproducible and modifiable analyses that are essential to the scientific enterprise. 

Computational sciences are then presented with a modern version of the reproducibility problem as computational research is largely inaccessible let alone reusable.
If research materials and outputs are shared, they are rarely disseminated in a readily modifiable or extensible format despite the great potential for distributing these resources digitally (e.g. Figshare, GitHub).
The ZOON R package aims to address these issues for Species Distribution Modelling by allowing users to encode SDM analyses as workflows comprised of independently executable modules.
The module-workflow structure enables scientists to more easily to create and share components of their analysis; and then access, modify, reuse and combine the components of others (see below and Figure 1). 

A lack of sharing reduces the capacity for innovation and validation in SDM, as well as its capacity to be a "self-correcting process" [@roysoc_report].
For example, a fundamental dispute over the ability of SDM to detect environmental associations (Beale et al. 2008, Araujou et al. 2009, Beale et al. 2009) was left unresolved as only the original publication shared their analytical code (Beale et al. 2008).
Without the resources being made open the critical differences in these analytical steps are assumed rather than proven. 

A second example comes from Elith et al. (2006), who reported a study of exceptional scope(“16 modelling methods over 226 species from 6 regions of the world”).
The modelling benchmarks that were reported have been highly influential, yet cannot be updated or modified when new modelling methods are produced as the code and data were not shared.
A third example comes from meta-analyses [] and methodological reviews [@Yackulic?], which rely on laborious searches through text publications in the absence of shared code and modelling outputs.
SDM then has limited abilities to derive truly informative methodological conclusions that can be acted upon. 

These three examples broadly illustrate why the data and models underpinning SDM need to become more repeatable, reproducible, accessible, modifiable and updateable.
A productive science relies on the accumulation of knowledge that can be reused and appropriated to advance understanding and resolve methodological issues.
Thus while many software are pitched as being ‘easy to use’ or ‘user friendly’, this is infrequently the case for the secondary user.

This paper introduces version 0.4-22 of the `ZOON` R package, which enables users to address issues of compatibility, accessibility, sharing and reproducibility for their SDM analyses.
Firstly, we describe the modular structure of ZOON workflows and demonstrate how completed workflows can be shared, evaluated and extended.
We then illustrate these concepts, and the potential contributions of ZOON to SDM research, by recreating and extending two recent SDM analyses using `zoon` workflows.
To assist users a number of tutorials are provided on the github repository and as vignettes distributed with zoon – including  guides on creating workflows and modules, as well as full technical details for module developers.

### Reproducible software projects

Most of the SDM users replying to a survey in 2015 (Ahmed et al. 2015) were using either MaxEnt [@Phillips2006] or R [@R] as the first choice software for their analyses (also see @Touchon2016). 
The software market is diverse, however, including a variety of software developed specifically for SDM (e.g. MaxEnt [@Phillips2006], the BioMod and BioMod2 [Thuiller2009] and dismo [@Hijmans?] packages for R, SDMtoolbox [@Brown2014], openModeller [@deSouzaMunoz2009], BioEnsembles [@Diniz-Filho2009], ModEco [@Guo2010]) and software that has been appropriated for SDM studies, such as statistical software (WinBugs and OpenBugs [@Lunn2000]) and other, more general ecological software  (Domain [Carpenter1993], Canoco [@TerBraak2002]). 

Of the SDM software for multi model comparisons (e.g. OpenModeller, BioEnsembles, ModEco), BioMod (as an open-source R package) has the greatest potential to be modified and re-appropriated.
In essence, BioMod is an SDM environment accessed from R, rather than necessarily being a SDM tool kit for R.
The R package called sdm [@Naimi2016] implements a variety of analytical functionality in a simplifying framework.
There is scope for incorporating new techniques but not necessarily a straight forward system for sharing those contributions.
The BIOVEL workflow environment [@DeGiovanni2015] is an exception in terms of reproducibility, sharing many of the general principles of ZOON.
Adoption of workflow environments (e.g. Kepler, Vis-Trails, Taverna) is low however, presumably due to the users’ unfamiliarity, and the requirement of embedding R within a workflow environment rather than embedding a workflow system within R.

Examples of subject specific frameworks are found within the Bioinformatics community, for example the Bioconductor infrastructure that uses R (Gentleman et al. 2004, https://www.bioconductor.org/), and BioJS, a repository for JavaScript components (Corpas et al. 2014, https://www.biojs.net/).
Both software aim to increase the compatibility and sharing of computational tools in order to increase the reuse and robustness of tools.
In terms of robustness, the CHASTE project for cardiac modelling (Mirams et al. 2013, http://www.cs.ox.ac.uk/chaste/) has developed into a model repository that offers community benchmarks for the last 50 years of cardiac modelling (Cooper et al. 2016).
These frameworks increase the access, usability and robustness of ultimately scientific research.
Of course, R users could point to the R environment itself, but the stringent requirements of R packages are likely to prevent much of the SDM community from sharing code in a discoverable or usable format.

### The ZOON project

The ZOON project takes a different approach, instead defining only the core architecture of workflows, and interfacing with an open repository of community-contributed software modules.
Users can create new modules for any modelling method and integrate it into another analysis. 

Similarly to programming languages such as R and Python, facilitating wider development of software will enable the whole research community to develop the software that it requires, rather than rely on the developers of the software they use to do it for them.
Unlike software packages for programming languages however, the ZOON project enforces a modular structure that ensures interoperability between contributed software modules.
This also helps to lower the bar for creating modules (each module is simply an R function with a set of required inputs and outputs) and facilitates automated testing of modules ot ensure software quality .

This paper introduces version 0.4-22 of the `zoon` R package, which can be dowloaded from CRAN and which encodes the majority of this functionality. First, we describe the modular structure of zoon workflows and demonstrate how completed `zoon` workflows can be shared, evaluated and extended.
Then we illustrate these concepts, and the contribution we hope `zoon` will make to SDM research, by recreating and extending two recent SDM analyses using `zoon` workflows.

# The `zoon` R package

The `zoon` R package encodes SDM analyses as a simple workflow of five key steps, with each step encoded as one or more software modules. Users combine these modules via a call to the `workflow` function, which executes each module in turn, before returning a `zoonWorkflow` object - a shareable, extensible and fully reproducible documentation of the SDM analysis. Figure \ref{fig:workflows} illustrates the structure of a `zoon` workflow comprising the five module types and illustrates how multiple modules may be combined at each stage. The module types and module combination options are described in more detail below.

<!— link to powerpoint doc: https://www.dropbox.com/s/afjpv8v6na9y56e/diagrams.pptx?dl=0-->

![The modular SDM structure encoded by a zoon workflow. A) Flow diagram representing the five module types and intermediate objects: df - an occurrence dataframe; ras - a RasterStack object; mod - a ZoonModel object; ? - any user-defined inputs or outputs. B) The flow diagram implied by chaining modules of the same type. C) The flow diagram implied by listing modules of the same type. Full details of module inputs and outputs, and the effects of listing and chaining each module type are given in the `zoon` vignette 'Building a module'. \label{fig:workflows}](./figs/diagrams.png)

####

#### Module types

The five module types, corresponding to the five key steps of an SDM anaysis are:

<!— needs work—>

- *Occurrence* - Usually presence-absence data or presence-only data, though abundance data is also used.
- *Covariates* - Predictor variables or covariates (typically environmental covariates) are required and the values of these covariates, at the locations of the occurrence data, must be extracted.
- *Process* - Processes  applied to the occurrence and covariate data. These processes include data cleaning, data thinning to account for spatial biases, feature selection using PCA or association tests and the splitting of data into training and test sets or cross validation folds.
- *Model* - Once the data has been suitable manipulated a model is fitted to estimate the relationships between the covariates and occurrence data. These models include simple statistical models such as GLMs as well as modern, flexible machine-learning methods such as MaxEnt and boosted regression trees.
- *Output* - The predictive power and goodness of fit must be assessed and the model parameters or response curves must be examined. The model is likely to be used to predict species occurrence, either in the vicinity of the occurrence data or elsewhere, or into the past or future.

#### Lists and Chains

<!— needs work—>

To combine multiple modules of the same type we provide the `Chain` command.
For occurrence and covariate modules, this command takes multiple modules and simply combines the data acquired by each module.
Chained process models are run sequentially.
For example if a user wants to generate background or pseudo absence data and then split the data into crossvalidation folds, modules implementing these two seperate process would be chained in that order.
Finally, chained output modules are simply all run seperately allowing the user to create multiple maps and summary figures, calcualte performance metrics and create other model outputs in one workflow.
Model modules cannot be chained.

Many SDM analyses require running similar workflow in parallel. 
In applied settings this may be running the same analysis for multiple species.
In methodological work, it is important to compare new methods or models to benchmarks.
To divide a `zoon` workflow into multiple parallel analyses the `list` function is used.
`list` can be used to split the analysis at any point: listing multiple occurrence modules together will run the workflow on different occurrence datasets while listing multiple model modules will fit different models to a shared, identical dataset but then apply output modules to each model separately.

The complete workflow, with multiple modules defined with the `Chain` function and parallel analyses being defined with the `list` function can be visualised by calling `plot` with the workflow object as it's argument.
This displays the names of all modules and the broader structure of the workflow as well as displaying which modules are not available in the online repository (Figure ).

### Inspecting, sharing and extending a `zoon` workflow


A `zoonWorkflow` objectStructure of workflow objects:

* code (call and modules used)
* output of each module; data, results and intermediate steps
* recording the session info and package and module versions

`zoon` workflows are re-runnable records of the data, code and results of an entire SDM analysis and can be easily reproduced, scrutinized and extended by the whole research community.
A zoon workflow returns an object of class `zoonWorkflow`.
This object contains all the information needed to reproduce the analysis.
This includes the original call, a record of the packages and modules used and their versions when the analysis was run as well as all the data and fitted models.

A workflow is reproducible, reusable and alterable by way of the functions `RerunWorkflow` and `ChangeWorkflow`.
Using `RerunWorkflow` a workflow object can be run from scratch or from part way though the analysis to avoid computationally slow steps.
`ChangeWorkflow` takes as it's arguments a workflow object and new selections for any or all of the workflow steps.
The workflow is then rerun from the first altered step.



Things you can do to workflows: 

* visualise the structure
* execute whole thing from scratch (grabs new data from web)
* execute from part way through

`zoon` provides functionality to quickly update an existing workflow object (even one created by someone else) by switching out modules.


Include a figure visualising the structure of the workflow object and how it can be adopted in the ChangeWorkflow function (corresponding to an example below?).

#### Building modules

Modules are simply R function definitions, with a set of required inputs and outputs.


As long as a valid module function is defined in an R session, it can be used in workflows in combination with modules downloaded from the ZOON module repository.
Whilst this makes it easy to develop modules locally, the strength of `zoon` comes from the ability to upload modules to the online repository so that others can access them.

To do this, it first is necessary to provide additional metadata including the module tile and description, module type, author name and definitions for any non-mandatory arguments.
`zoon` provides the function `BuildModule()` to facilitate entering this metadata and checking that it matches the module.

The `zoon` vignette *Building modules* provides a detailed tutorial for building modules of each type.



#### Additional features

##### Accessing module documentation

<!-- to do -->

##### Cross-validation

<!— needs work—>

`workflow` sensibly handles cross-validation.
Process modules are used to split the data into a training and test set or into multiple folds so that each fold in turn will be held back and used as the test for the predictive ability of any fitted models.
The same test/train splits will be used for each analysis if the multiple models or methods are being used so that model performance measures are comparable.
Furthermore, the test/train splits are saved along with the workflow so that other researchers adding to a workflow can make fair comparisons. 
After models are fitted to these subsets of the data, they are also fitted to the entire dataset as this will provide the best models for predictions from further new data.

##### Sharing workflows

`zoon` provides the `ZoonFigshare` function to facilitate sharing a completed workflow object from within R, via the free web platform [`figshare.com`](). `ZoonFigshare` takes a workflow object and some minimal metadata and uploads the workflow as an RData object to the user's figshare profile, for others to download, inspect and modify.



### Example Applications

We demonstrate the `zoon` R package by recreating two SDM analyses from published research articles as `zoon` workflows and showing how these workflow objects can easily be extended with new modules.
Workflow objects created by these analyses can be accessed at [http://figshare.com/articles/zoon_applications_paper_workflows](http://figshare.com/articles/zoon_applications_paper_workflows).
We encourage readers to download, interrogate and alter these workflows for themselves.
Full code and metadata for all of the modules used in the examples below can be found at [https://github.com/zoonproject/modules/R](https://github.com/zoonproject/modules/R)

#### Example 1. Modelling the potential distribution of nine-banded armadillo in the USA

@Feng2015 constructed a MaxEnt species distribution model for nine-banded armadillo in the USA using presence-only data on the species' current distribution, and the bioclim [@Hijmans2005] set of environmental correlates.
This model was then used to predict areas in the Americas which may be suitable for the species to become established. 

Such a model can be quickly and easily re-constructed as a `zoon` workflow using modules available in the ZOON module repository.
@Feng2015 used a combination of occurrence data from GBIF, and additional occurrence data manually collected from the published literature.
Unfortunately the latter data have not been made publically available, so here we use only data from GBIF.
If the the additional data had been made available it would be straightforward to incorporate them, for example using the `LocalOccurrenceData` module.

```{r loadzoon, eval = TRUE, echo = FALSE, cache = FALSE}
# Keep cache = FALSE. Not supposed to cache chunks with library()
#  I assume we want this not echoed.
library(zoon)

set.seed(1633)

```


```{r fengworkflow, eval = TRUE, dpi = 300, fig.show = "hide", results = "hide", message = FALSE, fig.height = 9, fig.width = 9}
Feng_Papes <- 
  workflow(occurrence = SpOcc('Dasypus novemcinctus',
                              extent = c(-130, -20, -60, 60)),
            covariate = Bioclim(extent = c(-130, -20, -60, 60),
                                layers = c(1:4, 6, 9, 10, 12, 15)),
              process = Chain(Clean, 
                              MESSMask,  
                              Background(n = 10000, bias = 200), 
                              Crossvalidate(k = 5)),
                model = MaxEnt,
               output = PrintMap(points = FALSE))
```

This module plots a static map of the predicted distribution, and the underlying data.
This map is shown in \ref{fig:feng_papes_plots}A.
The resulting workflow contains all the code required to re-run the workflow, the input data and the results of executing each module.
The object `Feng_Papes` could therefore be saved as a binary file and shared as a reproducible representation of this research.

Next, we update the workflow to produce an interactive map enabling anyone to inspect the data and predictions on a zoomable map, and to inspect the response curves of the fitted model. These outputs are shown in Figure This map is shown in \ref{fig:feng_papes_plots}, panels B and C.

```{r fengChangeWorkflow, eval = TRUE, fig.show = "hide", fig.height = 9}
Feng_Papes_Update <- 
  ChangeWorkflow(workflow = Feng_Papes,
                   output = Chain(InteractiveMap, 
                                  ResponseCurve(cov = 1)))
```

```{r combinefengupdate, eval = TRUE, echo = FALSE, dpi = 600, fig.show = "hide", fig.height = 3.2}
# combine the static and interactive maps and first effect plot

# load both images
r_map1 <- brick('figs/fengworkflow-1.png')
r_map <- brick('figs/interactive_map.png')
r_resp <- raster('figs/fengChangeWorkflow-1.png')
r_resp <- brick(r_resp, r_resp, r_resp)

# set up layout so that heights are the same for both panels
plot_height <- min(nrow(r_map1), nrow(r_map), nrow(r_resp))

# get rescaling factor for each
rescale_map1 <- plot_height / nrow(r_map1)
rescale_map <- plot_height / nrow(r_map)
rescale_resp <- plot_height / nrow(r_resp)

# get widths
width_map1 <- ncol(r_map1) * rescale_map1
width_map <- ncol(r_map) * rescale_map
width_resp <- ncol(r_resp) * rescale_resp

# set up layout (maximum total of 200 columns) with a gap in between
# mar won't work for these
gap <- 2
widths <- c(width_map1, width_map, width_resp)
widths <- round(widths * ((200 - (2 * gap)) / sum(widths)))
widths <- c(widths[1], gap, widths[2], gap, widths[3])
mat <- matrix(rep(1:5, widths),
              nrow = 1)
layout(mat)

# plot statis map
plotRGB(r_map1, maxpixels = Inf)

# add a panel letter
mtext(text = 'A',
        side = 3,
        line = -3,
        adj = 0)  

# gap
plot.new()

# plot interactive map
plotRGB(r_map, maxpixels = Inf)

# add a panel letter
mtext(text = 'B',
        side = 3,
        line = -3,
        adj = 0)  

# gap
plot.new()

# plot the response curve
plotRGB(r_resp, maxpixels = Inf,
        scale = max(maxValue(r_resp)))

# add a panel letter
mtext(text = 'C',
        side = 3,
        line = -3,
        adj = 0)  

```

![Outputs of the workflow objects `Feng_Papes` and `Feng_Papes_Update`. A) Map of the presence data (red), background data (black) and the MaxEnt relative probability of occurrence generated by the `PrintMap` module in the workflow `Feng_Papes`. White areas are masked due to being in the sea or outside the MESS mask. B) A screenshot of the interactive map produced by the `InteractiveMap` moduels in the workflow `Feng_Papes_Update`, displaying raw occurrence data and predicted distribution over a global map, allowing users to interactively explore their results. C) A response curve produced by the `ResponseCurve` module for the first covariate, bio1 in the workflow `Feng_Papes_Update`. \label{fig:feng_papes_plots}](./figs/combinefengupdate-1.png)

<!-- add the original and the response curves to this figure too -->



#### Example 2. Building a spatial thinning module

@Aiello-Lammens2015 proposed an approach for dealing with spatial sampling bias in presence-only data, by 'thinning' the presence data, and provide an R package `spThin` to implement their procedure [@spThin].
We can incorporate this approach in a workflow, by defining a simple *process* module that adapts the `zoon` data into the spThin format, uses this package to apply the algorithm, and converts data back into `zoon`'s expected format again:

```{r spthinmodule, eval = TRUE}
spThin <- function (.data, thin = 50) {
  
  # check these are presence-background data
  stopifnot(all(.data$df$type %in% c('presence', 'background')))
  
  # install & load the package
  zoon::GetPackage('spThin')
  
  # get dataframe & index to presence data
  df <- .data$df
  pres_idx <- which(df$type == 'presence')
  
  # prepare presence data subset and apply thinning
  sub_df <- data.frame(LAT = df$latitude[pres_idx],
                       LONG = df$longitude[pres_idx],
                       SPEC = NA)
  th <- thin(loc.data = sub_df,
             thin.par = thin,
             reps = 1,
             locs.thinned.list.return = TRUE,
             write.files = FALSE,
             write.log.file = FALSE)
  
  # get index to rows in sub_df, update the full dataset and return
  pres_keep_idx <- rownames(th[[1]])
  .data$df <- rbind(df[pres_idx[pres_keep_idx], ],
                    df[-pres_idx, ])
  return (.data)
}
```

To convert this code into a ZOON module, we need to write it to a standalone file (named `spThin.R`) with the metadata required to build the module documentation. The `zoon` function `BuildModule()` helps with this step, and can also run checks to make sure we got everything right:

```{r buildmodule, eval = TRUE, results = 'hide'} 
BuildModule(object = spThin,
            type = 'process',
            title = 'Spatial thinning of Presence-only Data',
            description = paste('Apply the stochastic spatial thinning',
                                'algorithm implemented in the spThin',
                                'package to presence data in a',
                                'presence-background dataset'),
            details = paste('Full details of the algorithm are available in',
                            'the open-access article by Aiello-Lammens',
                            'et al. (2015): dx.doi.org/10.1111/ecog.01132'),
            author = 'ZOON Developers',
            email = 'zoonproject@gmail.com',
            paras = list(thin = paste('Thinning parameter - the required',
                         'minimum distance (in kilometres) between points',
                         'after applying the thinning procedure')),
            dataType = 'presence-only',
            check = TRUE)
```

This module can now be shared so that others can use it in their ZOON workflows. The bexst way of sharing it is via the `zoon` modules repository. Modules can uploaded there, via the module submission system at [URL TBD].

#### Example 3. Evaluating MaxLike

@Royle2012 proposed a model for fitting SDMs to presence-only data, which they suggest is able to estimate the absolute (rather than relative) probability of species presence in each grid cell.
This model and the associated claims have been the subject of several research articles evaluating the model against both real-world and simulated datasets [@Fitzpatrick2013;@Merow2014;@Hastie2013;@Phillips2013].

In this example we illustrate how `zoon` can help contribute to discussions such as these by enabling rapid evaluation of a new method against a variety of datasets and test procedures.

For this example, it was necessary to create three new modules: two to load the presence/absence and presence-only Carolina Wren datasets from @Royle2012; and one to implement the MaxLike model.

<!--
Possible experiments:
* evaluating with different criteria (e.g. deviance or pseudo-r-squared which measure calibration capacity)
* fitting the same models with GBIF data and evaluating against the BBS PA data,
* fitting/evaluating on spatially-stratified holdout data
* fitting the models with other species
  (choose only one of these)
  -->

All of these have now been uploaded to the ZOON modules repository under the names `CarolinaWrenPO`, `CarolinaWrenPA`, `CarolinaWrenRasters` and `MaxLike`.

What the workflow looks like:

```{r merowworkflowtemp, eval = TRUE}
Merow_Silander <- workflow(
    occurrence = CarolinaWrenPO,
    covariate = CarolinaWrenRasters,
    process = Background(n = 10000),
    model = list(LogisticRegression, MaxEnt, MaxLike),
    output = Chain(PerformanceMeasures, PrintMap(points = FALSE)))
```


```{r merowworkflow, eval = FALSE}

Merow_Silander <- workflow(
    occurrence = CarolinaWrenPO,
    covariate = CarolinaWrenRasters,
    process = Chain(Transform(trans = 'square',
                                 replace = FALSE), 
                       Background(n = 10000)),
    model = list(LogisticRegression, MaxEnt, MaxLike),
    output = Chain(PerformanceMeasures, PointBiserialCorr, PrintMap(points = FALSE), Coefficients))
```

So we ran it again with disc-based spatial stratification:

```{r merowChangeWorkflow, eval = FALSE}
Merow_Silander_Spatial <- ChangeWorkflow(
    merow_and_silander,
    process = Chain(..., PartitionDisc))
```

### Future developments

Tutorials on how to create workflows and modules, as well as full technical details for module developers, are provided as vignettes distributed with `zoon`.

The `zoon` packages provides solutions to some of the technical barriers to the ultimate goal of more open, productive, reproducible SDM research.
Achieving this goal will also depend on overcoming the social hurdle of developing a common tool for comparing and quantitatively evaluating SDMs.

This will require achieving a critical mass of `zoon` modules, so that there is an incentive for species distribution modellers to use the software, and a critical mass of users, so that there is an incentive for SDM methods developers to encode their proposed methods as modules in the community repo.
At this point `zoon` will provide a common environment in which to evaluate SDM methods and enable the SDM research community to make evidence-based decisions about best practice and the goals for development of the field.

In order to facilitate this social aspect, future work will develop an online platform to enable exploring contributed modules and workflows.
This platform will also provide an online space to discuss, and openly evaluate, proposed best practices in SDM.
The zoon R package therefore represents a step towards a more reproducible ecosystem of SDM software. 

<!— point to github development site—>


### References


